\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles

### 2023-11-14 Manuscript/manuscript.tex
Format manuscript according Bio Adv template

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}
\author[1]{Boyi Guo}
\author[2]{Louise A. Huuki-Myers}
\author[3,4]{Melissa Grant-Peters}
\author[2]{Leonardo Collado-Torres}
\author[1,5,6,7]{Stephanie C. Hicks\thanks{Correspondence to \url{shicks19@jhu.edu}}}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA} 
\affil[2]{Lieber Institute for Brain Development, Johns Hopkins Medical Campus, Baltimore, MD, USA}
\affil[3]{Genetics and Genomic Medicine, Great Ormond Street Institute of Child Health, University College London, London, UK}
\affil[4]{Aligning Science Across Parkinson’s (ASAP) Collaborative Research Network, Chevy Chase, MD, USA}
\affil[5]{Department of Biomedical Engineering, Johns Hopkins School of Medicine, Baltimore, MD, USA}
\affil[6]{Center for Computational Biology, Johns Hopkins University, Baltimore, MD, USA}
\affil[7]{Malone Center for Engineering in Healthcare, Johns Hopkins University, MD, USA}

@@

Here, we focus on two types of visualizations in biomedical research, but note that the proposed ideas could be extended beyond these applications: (i) embedding visualizations, which project data into some low-dimensional embedding or mathematical space (e.g. Principal Components Analysis %(PCA)
\cite{hotelling_1933}, $t$-distributed Stochastic Neighbor Embedding ($t$-SNE) \cite{vandermaaten_2008}, or Uniform Manifold Approximation and Projection (UMAP) \cite{becht_2019}) and (ii) \textit{in situ} visualizations \cite{dries_2021, Lewis_2021, odonoghue_2021}, which aim to visualize molecules captured from \textit{in situ} imaging or sequencing technologies where \textit{in situ} refers to `in its original place'. Both of these visualizations represent data in a 2D space and are motivated by recent advances in experimental technologies that profile molecules, including DNA, RNA, and proteins, at a single-cell or spatial resolution \cite{kashima_2020, moffitt_2022}. Some most popular technologies include single-cell/nucleus RNA-sequencing (sc/snRNA-seq)
\cite{aldridge_2020} and \textit{in situ} spatially-resolved transcriptomics \cite{larsson_2021}. 
% 
% \vspace{-0.15in}

@@

To address these challenges, here we leverage the Gestalt (German for “unified whole”) principles \cite{todorovic_2008, palmer_1999} as a path to visualize multi-dimensional data in 2D visualizations. We focus on the two types of data visualizations previously introduced that are widely used in biomedical research: (i) embedding visualizations and (ii) \textit{in situ} visualizations. We provide an R package, \texttt{escheR}, implementing these ideas, which is built on the state-of-the-art data visualization framework \texttt{ggplot2} in the R programming language. Finally, we comment on how these ideas could be extended to other types of visualization in biomedical research.

@@

Gestalt principles \cite{todorovic_2008, palmer_1999} refer to a set of rules describing how humans perceive and interpret visual information and are commonly applied in art and designs. Developed in the 1920s by German psychologists Max Wertheimer, Kurt Koffka and Wolfgang Kohler, these principles help humans perceive a set of individual elements as a whole.

Here, we leverage the principles to be able to visualize multi-dimensional data in a unified 2D plot. Our approach is to use the state-of-art data visualization framework \texttt{ggplot2} \cite{ggplot2} following the Grammar of Graphics \cite{wilkinson_2012} and map individual variables to different aesthetics to simultaneously display disparate variables. Specifically, we apply the figure-ground segmentation \cite{peterson_2010} in displaying two variables: one variable (e.g. expression) can be plotted as color-filled circles, serving as the \textit{figure}; one variable (e.g. spatial domains) can be plotted as the backgrounds of the circles, creating a \textit{ground} for the figure. In practice, we use the combination of \texttt{color} and \texttt{fill="transparent"} to create the background layer and \texttt{fill} to create the figure layer. When necessary to display an additional layer for a third variable, \texttt{shape} can be used to add symbols such as cross (+) and asterisk (*) to highlight in the spatial map.

For adjacent circles with limited space between them to display the background color, we use an economic implementation, colored outlines for these circles (\textbf{Figure~\ref{fig:visual}C}), inspired by watercolor effect \cite{pinna_1987, pinna_2001}. Watercolor effect describes the phenomenon in visual perception that surface color arises from thin boundaries and hence is applied here to perceive the background color in tight space. Overall, the figure-ground segmentation creates two isolated layers in visual perception to display the two variables while maintaining the relative spatial relationship serving as a reference between the two. In addition, other fundamental principles \cite{todorovic_2008}, such as proximity, similarity, continuity, and closure, incentivize the brain to group elements and dimensions in the visualization, guaranteeing an integrative perception of the complex multi-dimensional spatial map.

@@

The application of the proposed framework is not limited to \textit{in situ} visualizations of spatially-resolved transcriptomics data. It is broadly applicable to data mapped to any 2-dimensional coordinate system to simultaneously display multiple variables. Such systems include euclidean space (including spatial coordinate as a special case) and data-driven embedding space, for example, UMAP and $t$-SNE. To demonstrate, we applied the proposed visualization to address the challenge of simultaneously displaying cluster membership and gene expression in a single-cell UMAP plot. To address the overplotting problem, previous work proposed to apply hexagonal binning strategy to display the gene expression \cite{freytag_2020}. Here, the color-coded convex hulls are used to annotate different clusters of cells (\textbf{Figure~\ref{fig:embedding}A}). However, the convex hulls create substantial overlapping areas, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. To improve the interpretability of the visualization, we replace the convex hulls with color-coded hexagons boundaries (\textbf{Figure~\ref{fig:embedding}B}) to avoid possible membership confusion. We note that our contribution to improving the visualization is easily implemented without any modification of \texttt{schex} as both are built upon the Grammar of Graphics \cite{wilkinson_2012} standard.

@@

While the proposed design principle set-up a solid foundation, the proposed package \texttt{escheR} can be further optimized for image-based spatially resolved data \cite{caicedo_2017}. For example, to address possible overplotting of densely arranged single cells in tight space, strategies such as binning or hatching plots \cite{patrick2023spatial} can be incorporated. In addition, the flexibility to represent data points as polygons will be provided to allow visualization of cell morphology.
% Combining with a binning strategy, similar to \texttt{schex} \cite{freytag_2020} to avoid possible overplotting \rr{or hatching plot proposed in \texttt{lisaClust} \cite{patrick2023spatial}}, the proposed visualization can be also applied to visualized image-based spatially resolved data , in addition to aforementioned spot-based spatially resolved data (\textbf{Figure~\ref{fig:visual}}).  \rr{\texttt{escheR} could also be applied with kernel gene-weighted density plots from \texttt{Nebulosa} \cite{alquicirahernandez_2021} and other ggplot2-based visualizations.}
% \rr{Independently, \texttt{lisaClust} \cite{patrick2023spatial} visualize cell types within spatial domains by generating hatching plots for immunofluorescence and mass cytometry imaging data.}

@@

\subsection*{Acknowledgments}
We would like to acknowledge Nicholas J. Eagles, Kristen R. Maynard, Mina Ryten, Leon Di Stefano, and Lukas M. Weber (appearing in alphabetic order of last name) for their helpful comments, feedback and suggestions on \texttt{escheR} functionality. Nicholas J. Eagles and Kristen R. Maynard are employed by the Lieber Institute for Brain Development; Mina Ryten is employed by University College London; Leon Di Stefano and Lukas M. Weber are from the Johns Hopkins Bloomberg School of Public Health, Department of Biostatistics.

@@

\begin{itemize}[nosep]
    \item Boyi Guo (ORCiD: \href{https://orcid.org/0000-0003-2950-2349}{0000-0003-2950-2349})
    \item Erik D. Nelson (ORCiD: \href{https://orcid.org/0000-0001-8477-0982}{0000-0001-8477-0982})
    \item Louise A. Huuki-Myers (ORCiD: \href{https://orcid.org/0000-0001-5148-3602}{0000-0001-5148-3602})
    \item Melissa Grant-Peters (ORCiD: \href{https://orcid.org/0000-0003-0585-0971}{0000-0003-0585-0971})
    \item Leonardo Collado-Torres (ORCiD: \href{https://orcid.org/0000-0003-2140-308X}{0000-0003-2140-308X})
    \item Stephanie C. Hicks (ORCiD: \href{https://orcid.org/0000-0002-7858-0231}{0000-0002-7858-0231})
\end{itemize}

>>>>

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}
\author[1]{Boyi Guo}
\author[2]{Louise A. Huuki-Myers}
\author[3,4]{Melissa Grant-Peters}
\author[1,2]{Leonardo Collado-Torres}
\author[1,5,6,7]{Stephanie C. Hicks\thanks{Correspondence to \url{shicks19@jhu.edu}}}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA} 
\affil[2]{Lieber Institute for Brain Development, Johns Hopkins Medical Campus, Baltimore, MD, USA}
\affil[3]{Genetics and Genomic Medicine, Great Ormond Street Institute of Child Health, University College London, London, UK}
\affil[4]{Aligning Science Across Parkinson’s (ASAP) Collaborative Research Network, Chevy Chase, MD, USA}
\affil[5]{Department of Biomedical Engineering, Johns Hopkins School of Medicine, Baltimore, MD, USA}
\affil[6]{Center for Computational Biology, Johns Hopkins University, Baltimore, MD, USA}
\affil[7]{Malone Center for Engineering in Healthcare, Johns Hopkins University, MD, USA}

@@

Here, we focus on two types of visualizations in biomedical research, but note that the proposed ideas could be extended beyond these applications: (i) embedding visualizations, which project data into some low-dimensional embedding or mathematical space (e.g. Principal Components Analysis , $t$-distributed Stochastic Neighbor Embedding ($t$-SNE), and Uniform Manifold Approximation and Projection (UMAP)) and (ii) \textit{in situ} visualizations \cite{dries_2021}, which aim to visualize molecules captured from \textit{in situ} imaging or sequencing technologies where \textit{in situ} refers to `in its original place'. Both of these visualizations represent data in a 2D space and are motivated by recent advances in experimental technologies that profile molecules, including DNA, RNA, and proteins, at a single-cell or spatial resolution% \cite{kashima_2020, moffitt_2022}
. Some most popular technologies include single-cell/nucleus RNA-sequencing (sc/snRNA-seq)
and \textit{in situ} spatially-resolved transcriptomics. \cite{vandereyken2023methods} 
% 
% \vspace{-0.15in}

@@

To address these challenges, here we leverage the Gestalt (German for “unified whole”) principles \cite{todorovic_2008} as a path to visualize multi-dimensional data in 2D visualizations. We focus on the two types of data visualizations previously introduced that are widely used in biomedical research: (i) embedding visualizations and (ii) \textit{in situ} visualizations. We provide an R package, \texttt{escheR}, implementing these ideas, which is built on the state-of-the-art data visualization framework \texttt{ggplot2} in the R programming language. Finally, we comment on how these ideas could be extended to other types of visualization in biomedical research.

@@

Gestalt principles \cite{todorovic_2008} refer to a set of rules describing how humans perceive and interpret visual information and are commonly applied in art and designs. Developed in the 1920s by German psychologists Max Wertheimer, Kurt Koffka and Wolfgang Kohler, these principles help humans perceive a set of individual elements as a whole.

Here, we leverage the principles to be able to visualize multi-dimensional data in a unified 2D plot. Our approach is to use the state-of-art data visualization framework \texttt{ggplot2} \cite{ggplot2} following the Grammar of Graphics \cite{wilkinson_2012} and map individual variables to different aesthetics to simultaneously display disparate variables. Specifically, we apply the figure-ground segmentation \cite{todorovic_2008} in displaying two variables: one variable (e.g. expression) can be plotted as color-filled circles, serving as the \textit{figure}; one variable (e.g. spatial domains) can be plotted as the backgrounds of the circles, creating a \textit{ground} for the figure. In practice, we use the combination of \texttt{color} and \texttt{fill="transparent"} to create the background layer and \texttt{fill} to create the figure layer. When necessary to display an additional layer for a third variable, \texttt{shape} can be used to add symbols such as cross (+) and asterisk (*) to highlight in the spatial map.

For adjacent circles with limited space between them to display the background color, we use an economic implementation, colored outlines for these circles (\textbf{Figure~\ref{fig:visual}C}), inspired by watercolor effect \cite{pinna_2001}. Watercolor effect describes the phenomenon in visual perception that surface color arises from thin boundaries and hence is applied here to perceive the background color in tight space. Overall, the figure-ground segmentation creates two isolated layers in visual perception to display the two variables while maintaining the relative spatial relationship serving as a reference between the two. In addition, other fundamental principles \cite{todorovic_2008}, such as proximity, similarity, continuity, and closure, incentivize the brain to group elements and dimensions in the visualization, guaranteeing an integrative perception of the complex multi-dimensional spatial map.

@@

The application of the proposed framework is not limited to \textit{in situ} visualizations of spatially-resolved transcriptomics data. It is broadly applicable to data mapped to any 2-dimensional coordinate system to simultaneously display multiple variables. Such systems include euclidean space (including spatial coordinate as a special case) and data-driven embedding space, for example, UMAP and $t$-SNE. To demonstrate, we applied the proposed visualization to address the challenge of simultaneously displaying cluster membership and gene expression in a single-cell UMAP plot. To address the overplotting problem, previous work proposed to apply hexagonal binning strategy to display the gene expression \cite{freytag_2020}. Here, the color-coded convex hulls are used to annotate different clusters of \rr{hexagons, where the cluster label of each hexagon is determined by the most popular label within each hexagon bin} (\textbf{Figure~\ref{fig:embedding}A}). However, the convex hulls create substantial overlapping areas, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. To improve the interpretability of the visualization, we replace the convex hulls with color-coded hexagons boundaries (\textbf{Figure~\ref{fig:embedding}B}) to avoid possible membership confusion. We note that our contribution to improving the visualization is easily implemented without any modification of \texttt{schex} as both are built upon the Grammar of Graphics \cite{wilkinson_2012} standard.

@@

While the proposed design principle set-up a solid foundation, the proposed package \texttt{escheR} can be further optimized for image-based spatially resolved data. For example, to address possible overplotting of densely arranged single cells in tight space, strategies such as hatching plots \cite{patrick2023spatial} can be incorporated. In addition, the flexibility to represent data points as polygons will be provided to allow visualization of cell morphology.
% Combining with a binning strategy, similar to \texttt{schex} \cite{freytag_2020} to avoid possible overplotting \rr{or hatching plot proposed in \texttt{lisaClust} \cite{patrick2023spatial}}, the proposed visualization can be also applied to visualized image-based spatially resolved data , in addition to aforementioned spot-based spatially resolved data (\textbf{Figure~\ref{fig:visual}}).  \rr{\texttt{escheR} could also be applied with kernel gene-weighted density plots from \texttt{Nebulosa} \cite{alquicirahernandez_2021} and other ggplot2-based visualizations.}
% \rr{Independently, \texttt{lisaClust} \cite{patrick2023spatial} visualize cell types within spatial domains by generating hatching plots for immunofluorescence and mass cytometry imaging data.}

@@

\subsection*{Acknowledgments}
We would like to acknowledge Nicholas J. Eagles, Kristen R. Maynard, Mina Ryten, Leon Di Stefano, and Lukas M. Weber (appearing in alphabetic order of last name) for their helpful comments, feedback and suggestions on \texttt{escheR} functionality. Nicholas J. Eagles and Kristen R. Maynard are employed by the Lieber Institute for Brain Development; Mina Ryten is employed by University College London; Leon Di Stefano is from the Johns Hopkins Bloomberg School of Public Health, Department of Biostatistics; Lukas M. Weber is in the Department of Biostatistics at Boston University School of Public Health.

@@

\begin{itemize}[nosep]
    \item Boyi Guo (ORCiD: \href{https://orcid.org/0000-0003-2950-2349}{0000-0003-2950-2349})
    \item Louise A. Huuki-Myers (ORCiD: \href{https://orcid.org/0000-0001-5148-3602}{0000-0001-5148-3602})
    \item Melissa Grant-Peters (ORCiD: \href{https://orcid.org/0000-0003-0585-0971}{0000-0003-0585-0971})
    \item Leonardo Collado-Torres (ORCiD: \href{https://orcid.org/0000-0003-2140-308X}{0000-0003-2140-308X})
    \item Stephanie C. Hicks (ORCiD: \href{https://orcid.org/0000-0002-7858-0231}{0000-0002-7858-0231})
\end{itemize}
